Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-04-20 doc/invt.tex
small edits to invt.tex

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:y2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_split}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where we have approximated $\theta_k(t)$
as white noise:
%
\begin{equation}
  \kappa_k(t)
  \equiv \langle \theta_k(0) \, \theta_k(t) \rangle
  \approx \Gamma_k \, \delta(t)
  .
  \notag
  %\label{eq:kappa_delta}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}

For the theory to be applicable to
a general sampling process,
we can map the histogram fluctuation
to the equivalent white noise, with
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  %= \sum_{t = -\infty}^\infty \left\langle \theta_k(0) \, \theta_k(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
In this way,
the underlying sampling process
affects the error only through the
few $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to long-range fluctation
in a local sampling process,
especially for complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.
%
%\begin{equation}
%  \Err_A(T)
%  =
%  \int_0^T \int_0^T
%  \sum_{k = 2}^n
%    \dot u_k\bigl( q(t) \bigr) \,
%    \dot u_k\bigl( q(t') \bigr) \,
%    \kappa_k(t - t') \, dt \, dt'
%  ,
%  \notag
%  %\label{eq:error_asym0}
%\end{equation}
%%
%with
%$u_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}$,
%and
%$\kappa_k(t - t') \equiv \left\langle \theta_k(t) \, \theta_k(t') \right\rangle$.
%
%\begin{align*}
%  u_k(q')
%  &\equiv
%  e^{\lambda_k \, [q' - q(T)]},
%  \\
%  \kappa_k(t - t')
%  &\equiv
%  \left\langle
%    \theta_k(t) \, \theta_k(t')
%  \right\rangle.
%\end{align*}

@@

We will minimize the error under a fixed value of $q(T)$.
%
This allows us to focus on the asymptotic part
for the residual error
depends on $\alpha(t)$ only through
the integral, $q(T) = \int_0^T \alpha(t) \, dt$.
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t )
  .
\notag
\end{align}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle.
%
So the above Lagrangian contains only the kinetic energy,
and, therefore, describes a free particle
carrying a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved.
%
But in this case the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
coincides with the Lagrangian.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  \left(
    \frac{C}{T}
  \right)^2
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive constant, $C$.
%
This serves
as a generalization of Eq. \eqref{eq:dudt_const}.
%
By integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
where $C$ is determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $\Err_A(T)$,
in the above,
the value of $q(T)$ can be further adjusted
to minimize the total error, $\Err(T) = \Err_R(T) + \Err_A(T)$
(cf. Sec. \ref{sec:optinitalpha}).

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define a mass function as
%
\begin{equation}
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
\label{eq:mass_func}
\end{equation}
%
and a normalized mass distribution as
%
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%

@@

    % label for the shaded area
    \node[] at (1.1, 0.5) {$\dfrac{t}{T}$};

@@

The above characterization allows us to determine
the optimal initial updating magnitude.
%
Let us now consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude, $a_0$.
%
For these simulations,
we can adjust $q(T)$
to minimize the total error.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (a_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial Q_c
  }
  &=
  -a_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial Q_c
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}%
By Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we can rewrite the condition as
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is always half of the equilibrium value.
%
This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, *wang2001pre}.

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
An example is the Gaussian updating scheme
used in metadynamics.

@@

%It is natural to first introduce homogeneous updating schemes
For a periodic variable\cite{dama2014},
the updating matrix, $\mathbf w$,
assumes the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $|l| > b$.
%
We will further impose the symmetry
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
These numbers constitute an \emph{updating kernel}\cite{bussi2006}.
%
As shown in Appendix \ref{sec:wband_eig},
the multiplication of $\mathbf w$
is equivalent to a convolution with the kernel, and
the eigenvalues of $\mathbf w$ are
the Fourier or cosine transform of the updating kernel:
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2 \,
  \sum_{ l = 1 }^b
  \mu_l
  \cos
  \frac{ (k - 1) \, l \, 2 \, \pi }
       {            n             }
  ,
  \label{eq:wband_eigenvalue_pbc}
\end{equation}
%
with a two-fold degeneracy,
$\lambda_{n - k + 2} = \lambda_k$.

@@

Note that the two reflective mirrors effectively
define a periodic variable of period $2 \, n$.
Thus, the eigenvalues can be borrowed from
Eq. \eqref{eq:wband_eigenvalue_pbc},
with the period, $n$, doubled:
%
\begin{align}
  \lambda_k
  &=
  \mu_0
  +
  2
  \sum_{l = 1}^b
    \mu_l
    \cos \frac{(k - 1)  \, l \, 2 \, \pi}{ 2 \, n }
  .
\label{eq:wband_eigenvalue_refl}
\end{align}
%
We may unify Eqs. \eqref{eq:wband_eigenvalue_refl}
and \eqref{eq:wband_eigenvalue_pbc} as
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2
  \sum_{ l = 1 }^b
    \mu_l \,
    \cos \frac{ (k - 1) \, l \, 2 \, \pi }
              {            g \, n        }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where the $g = 1$ for the periodic case,
or $2$ otherwise.
%
In both cases,
the eigenvalues, $\lambda_k$'s,
are the cosine transform of
the updating kernel.

@@

\item
  Compute the error $\Err(T)$ from
Eqs. \eqref{eq:error_split},
  \eqref{eq:error_res1},
  and
  \eqref{eq:error_asym2}.

@@

By including the residual error,
we determine the optimal $t_0$
that minimizes the total error.
%
According to Eq. \eqref{eq:half_alpha0}, we have
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  .
\label{eq:t0_sinc}
\end{equation}
%
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:y2_eql}.
%
From Eqs. \eqref{eq:error_split},
\eqref{eq:error_res1},
\eqref{eq:tk0},
and
\eqref{eq:error_asym_sinc},
we get
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
which serves as a generalization of
the result from the single-bin scheme,
Eq. \eqref{eq:Emin_singlebin}
[one can show that $\sum_{k=2}^n \Gamma_k = \sum_{i=1}^n \rho_i \, G_i = \Gamma$,
following the steps similar to those leading to Eq. \eqref{eq:y2_sum}].
%
\note{The derivation.
From Eq. \eqref{eq:eig_orthonormal_cols}, we have
$\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle
=\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle$,
so
\begin{align*}
\sum_{i=1}^n \rho_i G_i
&=\sum_{t=-\infty}^\infty\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle \\
&=\sum_{t=-\infty}^\infty\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle
=\sum_{k=2}^n \Gamma_k.
\end{align*}
}
%
This error is roughly proportional to the inverse simulation length,
as one would expect from an equilibrium FDS simulation.
%In Appendix \ref{sec:equilerr},
%we show that the above error tends
%to the error of the ideal equilibrium FDS simulation
%in the asymptotic limit.

@@

For a non-periodic variable,
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that the matrix multiplication is still equivalent to
a convolution as Eq. \eqref{eq:wmul_to_convol}.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
We note that Eq. \eqref{eq:phi_refl}
defines a periodic variable of period $2 \, n$
with a reflective symmetry around $i = 1/2$.
%
Accordingly, the eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
share the same periodicity,
and are even functions about the axis, $i = 1/2$:
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, 2 \, \pi}
            {                    2 \, n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
and the eigenvalues are given by
  Eq. \eqref{eq:wband_eigenvalue_refl}.

\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue_refl}.
}%
%

@@

We will derive here an upper bound of the asymptotic error
that can be associated to updating schemes
with a set of well-separated eigenvalues,
in parallel to the lower bound reached by
the single-bin scheme with a set of identical eigenvalues
(cf. Sec. \ref{sec:optWL}).

>>>>

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:y2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  u_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \langle \theta_k(0) \, \theta_k(t) \rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty \left\langle \theta_k(0) \, \theta_k(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
Equation \eqref{eq:Gamma_sum}
essentially maps the histogram fluctuation, $\theta_k(t)$,
to the equivalent white noise
so that the theory can be applicable to
a general sampling process.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to
persistent long-range motion
in sampling complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.
%
%\begin{equation}
%  \Err_A(T)
%  =
%  \int_0^T \int_0^T
%  \sum_{k = 2}^n
%    \dot u_k\bigl( q(t) \bigr) \,
%    \dot u_k\bigl( q(t') \bigr) \,
%    \kappa_k(t - t') \, dt \, dt'
%  ,
%  \notag
%  %\label{eq:error_asym0}
%\end{equation}
%%
%with
%and
%$\kappa_k(t - t') \equiv \left\langle \theta_k(t) \, \theta_k(t') \right\rangle$.
%
%\begin{align*}
%  u_k(q')
%  &\equiv
%  e^{\lambda_k \, [q' - q(T)]},
%  \\
%  \kappa_k(t - t')
%  &\equiv
%  \left\langle
%    \theta_k(t) \, \theta_k(t')
%  \right\rangle.
%\end{align*}

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 2 }^n
  %  \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(Q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
\label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,
and we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$
%
This serves
as a generalization of Eq. \eqref{eq:dudt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(Q)
    \;
    d Q
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    M( Q )
    \;
    d Q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define
a normalized mass distribution as
%
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%

@@

    % label for the shaded area
    \node[] at (1.1, 0.40) {$\dfrac{t}{T}$};

@@

The above characterization allows us to determine
the optimal initial updating magnitude,
or, equivalently, the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
We will consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude, $a_0$.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (a_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial Q_c
  }
  &=
  -a_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial Q_c
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}%
By Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we can rewrite the condition as
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is always half of the equilibrium value.
%
This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, *wang2001pre}.

@@

Many commonly-used updating schemes,
such as the Gaussian updating scheme
used in metadynamics,
can be characterized by a rigid window
centered on the current bin.
%
For these translationally invariant or homogeneous
schemes, the eigenvalues can be readily found
from the shape of the window function.
%

@@

%It is natural to first introduce homogeneous updating schemes
For a periodic variable\cite{dama2014},
the updating matrix, $\mathbf w$,
assumes the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where the numbers,
$\mu_{-b}, \dots, \mu_b$, ($b \le n/2$)
characterizing the rigid window
will be referred to as an \emph{updating kernel}\cite{bussi2006}.
%
To satisfy Eq. \eqref{eq:w_sumj},
we impose the normalization condition
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
with $\mu_l = 0$ for $|l| > b$.
%
If $b = n/2$, then $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
For simplicity, we also impose the symmetry
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
As shown in Appendix \ref{sec:wband_eig},
the multiplication of $\mathbf w$
is equivalent to a convolution with the kernel, and
the eigenvalues of $\mathbf w$ are
the Fourier or cosine transform of the updating kernel:
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2 \,
  \sum_{ l = 1 }^b
  \mu_l
  \cos
  \frac{ (k - 1) \, l \, 2 \, \pi }
       {            n             }
  ,
  \label{eq:wband_eigenvalue_pbc}
\end{equation}
%
with a two-fold degeneracy,
$\lambda_{n - k + 2} = \lambda_k$.

@@

Note that the two reflective mirrors effectively
define a periodic variable of period $2 \, n$.
Thus, the eigenvalues can be borrowed from
Eq. \eqref{eq:wband_eigenvalue_pbc},
with a doubled period:
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2
  \sum_{ l = 1 }^b
    \mu_l \,
    \cos \frac{ (k - 1) \, l \, 2 \, \pi }
              {            g \, n        }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where $g = 1$ or $2$
in the periodic or non-periodic case, respectively.

@@

\item
  Compute the total error $\Err(T)$ from
Eqs. \eqref{eq:error_tot},
  \eqref{eq:error_res1},
  and
  \eqref{eq:error_asym2}.

@@

By including the residual error,
we determine the optimal $t_0$
that minimizes the total error.
%
According to Eq. \eqref{eq:half_alpha0}, we have
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  .
\label{eq:t0_sinc}
\end{equation}
%
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:y2_eql}.
%
From Eqs. \eqref{eq:error_tot},
\eqref{eq:error_res1},
\eqref{eq:tk0},
and
\eqref{eq:error_asym_sinc},
we get
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
which serves as a generalization of
the result from the single-bin scheme,
Eq. \eqref{eq:Emin_singlebin}
[one can show that $\sum_{k=2}^n \Gamma_k = \sum_{i=1}^n \rho_i \, G_i = \Gamma$,
following the steps similar to those leading to Eq. \eqref{eq:y2_sum}].
%
\note{The derivation.
From Eq. \eqref{eq:eig_orthonormal_cols}, we have
$\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle
=\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle$,
so
\begin{align*}
\sum_{i=1}^n \rho_i G_i
&=\sum_{t=-\infty}^\infty\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle \\
&=\sum_{t=-\infty}^\infty\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle
=\sum_{k=2}^n \Gamma_k.
\end{align*}
}
%
This error is roughly proportional to the inverse simulation length,
as one would expect from an equilibrium FDS simulation.
%In Appendix \ref{sec:equilerr},
%we show that the above error tends
%to the error of the ideal equilibrium FDS simulation
%in the asymptotic limit.

@@

For a non-periodic variable,
we define the out-of-boundary values
of $\phi_i$ as
%
\begin{equation}
  \phi_i
  =
  \begin{dcases}
    \phi_{ 1 - i }           & \mathrm{for \;} i \le 0, \\
    \phi_{ 2 \, n + 1 - i }  & \mathrm{for \;} i > n,
  \end{dcases}
\label{eq:phi_refl}
\end{equation}
%
such that the matrix multiplication is still equivalent to
a convolution as Eq. \eqref{eq:wmul_to_convol}.
%
\note{The derivation of Eq. \eqref{eq:wmul_to_convol}
  in this case is similar,
  $$
  \begin{aligned}
    \sum_{j = 1}^n w_{ij} \, \phi_j
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 1} \, \phi_j
    +
    \sum_{j = 1}^n
      \mu_{i + j - 2 \, n - 1} \, \phi_j
    \\
    &=
    \sum_{j = 1}^n
      \mu_{i - j} \, \phi_j
    +
    \sum_{l = 1 - n}^0
      \mu_{i - l} \, \phi_l
    +
    \sum_{l = n + 1}^{ 2 \, n }
      \mu_{i - l} \, \phi_l
    \\
    &=
    \sum_{j = 1 - n}^{ 2 \, n}
      \mu_{i - j} \, \phi_j.
  \end{aligned}
  $$
}%
We note that Eq. \eqref{eq:phi_refl}
defines a periodic variable of period $2 \, n$
with a reflective symmetry around $i = 1/2$.
%
Accordingly, the eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
share the same periodicity,
and are even functions about the axis, $i = 1/2$:
%
\begin{equation}
  \phi^{(k)}_i
  =
  \phi_{k i}
  =
  \frac{ \sqrt{ 2 - \delta_{k, 1} } }
       {             n              }
  \cos \frac{ ( k - 1) \, \left( i - \frac 1 2 \right) \, 2 \, \pi}
            {                    2 \, n                           }
  ,
\notag
%\label{eq:wband_eigenvector_refl}
\end{equation}
%
and the eigenvalues are given by
  Eq. \eqref{eq:wband_eigenvalue}
  with $g = 2$.

\note{Derivation.
  First consider the unnormalized eigenvectors,
  $$
  \Phi^{(k)}_i
  =
  \cos \frac{ \left( i - \frac 1 2 \right) (k - 1) \, \pi}{n},
  $$
  which satisfies Eq. \eqref{eq:phi_refl}.
  %
  So, by Eq. \eqref{eq:wmul_to_convol},
  $$
  \begin{aligned}
  \sum_{i = 1}^n
    \Phi^{(k)}_i \, w_{ij}
  &=
  \sum_{l = -b}^b
    \Phi^{(k)}_{j - l} \, \mu_l
  \\
  &=
    \Phi^{(k)}_j \, \mu_0
  + \sum_{l=0}^{b}
    \mu_l \,
    \left[
      \Phi^{(k)}_{j-l}
      +
      \Phi^{(k)}_{j+l}
    \right]
  \\
  &= \Phi^{(k)}_j \, \lambda_k,
  \end{aligned}
  $$
  with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue}
  with $g = 2$.
}%
%

@@

We will derive here an upper bound of the asymptotic error
that can be associated to updating schemes
with a set of well-separated eigenvalues,
in parallel to the lower bound reached by
the single-bin scheme with a set of identical eigenvalues
(cf. Sec. \ref{sec:optWL}).
%
For example, the smaller eigenvalues of the Gaussian updating scheme
are well separated because of the rapid exponential decay.
